Combinatorial interpretations of~holonomic recurrences

### 2021-09-15 article/main.tex
Fix biblatex margins
Last fixes

\RequirePackage[maxbibnames=99]{biblatex}
\addbibresource{bibliography.bib}

@@

The method itself has proven to be helpful in order to divide the problem into smaller parts and find some corner cases. Unfortunately, it does not give any immediate interpretation for applying the inductive step.
\todo{fill the gaps}
\todo{add reconstruction images}

@@

\begin{definition}
    The set of lambda terms \(\Lambda\) is defined recursively:
    \begin{itemize}
        \item if \(x\) is a~variable, then \(x \in \Lambda\)
        \item if \(M, N \in \Lambda\), then \((M N) \in \Lambda\) (it can be interpreted as an application of a~function \(M\) to an~argument \(N\))
        \item if \(x\) is a~variable and \(M \in \Lambda\), then \((\lambda x. M) \in \Lambda\) (it can be interpreted as an~abstraction, i.e.~creating a~function with an~argument \(x\) and binding each occurrence of \(x\) in \(M\) to this argument. We say that a variable \(x\) is abstracted by this lambda-abstraction)
    \end{itemize}
\end{definition}

This definition resembles tree-based structure of lambda terms, where:
\begin{itemize}
    \item lambda abstraction is represented as an unary node holding name of lambda-abstracted variable;
    \item application is represented as a binary node;
    \item variable is represented as a leaf holding name of the variable.
\end{itemize}

@@

The definition using named variables is quite popular, but it introduces several problems. One of them is a~problem of a potential collision of lambda-abstracted variable names and difficulty of finding to which lambda abstraction the variable is bound (or is it even a~free variable, i.e.~not bound). To address that, we can use de Bruijn indices.

\begin{definition}
    Set of lambda terms \(\LambdadB\) is defined recursively:
    \begin{itemize}
        \item if \(n \in \N\), then \(n \in \LambdadB\) (\(n\) is a~variable bound to a lambda abstraction \(n\) levels above, i.e.~if you traverse the tree of the lambda term upwards, we have to ignore \(n\) lambda-abstraction nodes and the \((n+1)\)-st one is the one binding the variable. If we would have to go beyond the root of the tree, we say that the variable is free, i.e.~not bound)
        \item if \(M, N \in \LambdadB\), then \((M N) \in \LambdadB\) (application)
        \item if \(M \in \LambdadB\), then \((\lambda M) \in \LambdadB\) (abstraction)
    \end{itemize}
\end{definition}

@@

As we have our structure defined, let's create a generating function that will keep track of more than just the number of nodes, but also counts the number of each kind of node in the tree. The generating function \(\gf{T}(z, o, l, a, s)\) will be parameterized then using the following arguments:
\begin{itemize}
    \item \(z\) marking all actual nodes
    \item \(o\) marking variable nodes
    \item \(l\) marking lambda abstraction nodes
    \item \(a\) marking application nodes
    \item \(s\) marking successors of variable nodes
\end{itemize}

@@

\begin{center}
    \begin{minipage}[t]{.2\textwidth}
        \begin{center}
            \(\gf{T}\)\\
            \includesvg[scale=0.5]{lambda__def__1}%
        \end{center}
    \end{minipage}%
    \begin{minipage}[t]{.05\textwidth}
        \begin{center}
            \(=\)\\
        \end{center}
    \end{minipage}%
    \begin{minipage}[t]{.2\textwidth}
        \begin{center}
            \(z a~\gf{T}^2\)\\
            \includesvg[scale=0.5]{lambda__def__2}%
        \end{center}
    \end{minipage}%
    \begin{minipage}[t]{.05\textwidth}
        \begin{center}
            \(+\)\\
        \end{center}
    \end{minipage}%
    \begin{minipage}[t]{.2\textwidth}
        \begin{center}
            \(z l \gf{T}\)\\
            \includesvg[scale=0.5]{lambda__def__3}%
        \end{center}
    \end{minipage}%
    \begin{minipage}[t]{.05\textwidth}
        \begin{center}
            \(+\)\\
        \end{center}
    \end{minipage}%
    \begin{minipage}[t]{.2\textwidth}
        \begin{center}
            \(\gf{S}\)\\
            \includesvg[scale=0.5]{lambda__def__4}%
        \end{center}
    \end{minipage}%
\end{center}

@@

If we treat \(z\) as the only variable and make \(o\), \(l\), \(a\), and \(s\) constants, we can use a~little bit of Maple magic~\cite{gfun} to achieve the following holonomic recurrence.

@@

\[\begin{array}{rl}
        0 =& (n - 4) t_{n - 4}\\
        +& t_{n - 3}\\
        +& (2 n - 1) t_{n - 2}\\
        +& (-4 n + 1) t_{n - 1}\\
        +& (1 + n) t_{n}
\end{array}.\]

>>>>

\RequirePackage[maxbibnames=99]{biblatex}
\addbibresource{bibliography.bib}
\appto{\bibsetup}{\sloppy}
% \appto{\bibsetup}{\raggedright}

@@

The method itself has proven to be helpful in order to divide the problem into smaller parts and to find some corner cases. Unfortunately, it does not give any immediate interpretation for applying the inductive step.
\todo{fill the gaps}
\todo{add reconstruction images}

@@

\begin{definition}
    The set of lambda terms \(\Lambda\) is defined recursively:
    \begin{itemize}
        \item if \(x\) is a~variable, then \(x \in \Lambda\);
        \item if \(M, N \in \Lambda\), then \((M N) \in \Lambda\) (it can be interpreted as an application of a~function \(M\) to an~argument \(N\));
        \item if \(x\) is a~variable and \(M \in \Lambda\), then \((\lambda x. M) \in \Lambda\) (it can be interpreted as an~abstraction, i.e.~creating a~function with an~argument \(x\) and binding each occurrence of \(x\) in \(M\) to this argument. We say that a variable \(x\) is abstracted by this lambda-abstraction).
    \end{itemize}
\end{definition}

This definition resembles a tree-based structure of lambda terms, where:
\begin{itemize}
    \item lambda abstraction is represented as an unary node holding name of lambda-abstracted variable;
    \item application is represented as a binary node;
    \item variable is represented as a leaf holding name of the variable.
\end{itemize}

@@

The definition using named variables is quite popular, but it introduces several problems. One of them is a~problem of a potential collision of lambda-abstracted variable names and difficulty of deciding to which lambda abstraction the variable is bound (or is it even a~free variable, i.e.~not bound). To address that, we can use de Bruijn indices.

\begin{definition}
    The set \(\LambdadB\) of lambda terms in de Bruijn notation is defined recursively:
    \begin{itemize}
        \item if \(n \in \N\), then \(n \in \LambdadB\) (\(n\) is a~variable bound to a lambda abstraction \(n\) levels above, i.e.~if we traverse the tree of the lambda term upwards, we have to ignore \(n\) lambda-abstraction nodes and the \((n+1)\)-st one is the one binding the variable. If we have to go beyond the root of the tree, we say that the variable is free, i.e.~not bound);
        \item if \(M, N \in \LambdadB\), then \((M N) \in \LambdadB\) (application);
        \item if \(M \in \LambdadB\), then \((\lambda M) \in \LambdadB\) (abstraction).
    \end{itemize}
\end{definition}

@@

As we have our structure defined, let's create a generating function that keeps track of more than just the number of nodes, but also counts the number of each kind of node in the tree. The generating function \(\gf{T}(z, o, l, a, s)\) will be parameterized then using the following arguments:
\begin{itemize}
    \item \(z\) marking all actual nodes,
    \item \(o\) marking variable nodes,
    \item \(l\) marking lambda abstraction nodes,
    \item \(a\) marking application nodes,
    \item \(s\) marking successors of variable nodes.
\end{itemize}

@@

\begin{center}
    \begin{minipage}[t]{.2\textwidth}
        \begin{center}
            \(\gf{T}\)\\
            \includesvg[scale=0.5]{lambda__def__1}%
        \end{center}
    \end{minipage}%
    \begin{minipage}[t]{.05\textwidth}
        \begin{center}
            \(=\)\\
        \end{center}
    \end{minipage}%
    \begin{minipage}[t]{.2\textwidth}
        \begin{center}
            \(z a~\gf{T}^2\)\\
            \includesvg[scale=0.5]{lambda__def__2}%
        \end{center}
    \end{minipage}%
    \begin{minipage}[t]{.05\textwidth}
        \begin{center}
            \(+\)\\
        \end{center}
    \end{minipage}%
    \begin{minipage}[t]{.2\textwidth}
        \begin{center}
            \(z l \gf{T}\)\\
            \includesvg[scale=0.5]{lambda__def__3}%
        \end{center}
    \end{minipage}%
    \begin{minipage}[t]{.05\textwidth}
        \begin{center}
            \(+\)\\
        \end{center}
    \end{minipage}%
    \begin{minipage}[t]{.2\textwidth}
        \begin{center}
            \(\gf{V}\)\\
            \includesvg[scale=0.5]{lambda__def__4}%
        \end{center}
    \end{minipage}%
\end{center}

@@

If we treat \(z\) as the only variable and make \(o\), \(l\), \(a\), and \(s\) constants, we can use a~little bit of Maple magic~\cite{gfun} to achieve the following holonomic recurrence:

\[\begin{array}{rl}
        0 =& (n - 4) l^2 s^2 t_{n - 4}\\
        +& ((4 n - 10) a o s - (2 n - 6) l^2 s - (2 n - 5) l s^2) t_{n - 3}\\
        +& ((-4 n + 8) a o + (n - 2) l^2 + (4 n - 6) l s + (n - 1) s^2) t_{n - 2}\\
        +& ((-2 n + 1) l - 2 n s) t_{n - 1}\\
        +& (n + 1) t_{n}.
\end{array}\]

@@

\[\begin{array}{rl}
        0 =& (n - 4) t_{n - 4}\\
        +& t_{n - 3}\\
        +& (2 n - 1) t_{n - 2}\\
        +& (-4 n + 1) t_{n - 1}\\
        +& (1 + n) t_{n}.
\end{array}\]
